Natural Selection is Unlikely to Explain Why Species Get a Thin Slice of $\pi$

### 2021-06-01 manuscript/manuscript.tex
minor update
minor changes

\title{Why do species get a thin slice of $\pi$? Revisiting Lewontin's Paradox of Variation}

@@

A major impediment in quantifying Lewontin's Paradox has been estimating census
population sizes across many taxa, especially for extremely abundant,
cosmopolitan species that define the upper limit of ranges. Previous work has
surveyed the literature for census size estimates
\parencite{Nei1984-zi,Soule1976-he,Frankham1996-yb}, or used range, body size,
or qualitative categories as proxies for census size
\parencite{Corbett-Detig2015-gt,Leffler2012-zj}. To quantify the relationship
between genomic estimates of diversity and census population sizes, I first
approximate census population sizes for 172 metazoan taxa (Figure
\ref{fig:figure-nc}). My approach predicts population densities from body sizes
using a previously-observed linear relationship that holds across metazoans
(Supplementary Figure \ref{suppfig:damuth};
\cite{Damuth1981-st,Damuth1987-sg}). Then, from geographic occurrence data, I
estimate range sizes.  Finally, I estimate population size as the product of
these predicted densities and range estimates (see Methods:
\Newnameref{sec:methods-popsize}). Note that the relationship between
population density and body size is driven by energy budgets, and thus reflects
macroecological equilibria \parencite{Damuth1987-sg}. Consequently, population
sizes are underestimated for taxa like humans and their domesticated species,
and overestimated for species with anthropogenically reduced densities or
fragmented ranges. For example, the population size of \emph{Lynx lynx} is
likely around 50,000 \parencite{Iucn2020-ap} which is around two orders of
magnitude smaller than my estimate. Additionally, the range size estimates do
not consider whether an area has unsuitable habitat, and thus may be
overestimated for species with particular niches or patchy habitats. While
these methods to estimate census size are crude and approximate, they can be
efficiently calculated for numerous taxa and are sufficient to estimate the
scale of Lewontin's Paradox (see Supplementary Information
\Newnameref{supp:popsize-valid} for more on validation based on biomass and
other approaches).

@@

To determine which ecological or evolutionary processes could decouple
diversity from census population size, we first need to quantify this
relationship across a wide variety of taxa. Previous work has found there is a
significant relationship between heterozygosity and the logarithm of population
size or range size, but these studies relied on heterozygosity measured from
allozyme data
\parencite{Soule1976-he,Frankham1996-yb,Nei1984-zi,Leffler2012-zj}. Here, I
confirm these findings using pairwise diversity estimates from genomic sequence
data and the estimated census sizes (Figure \ref{fig:figure-1}). The pairwise
diversity estimates are from three sources: \textcite{Leffler2012-zj},
\textcite{Corbett-Detig2015-gt}, and \textcite{Romiguier2014-bp}, and are
predominantly from either synonymous or non-coding DNA (see Methods:
\Newnameref{sec:methods-diversity}). Overall, an ordinary least squares (OLS)
relationship on a log-log scale fits the data well (Figure \ref{fig:figure-1}).
The OLS slope estimate is significant and implies a 13\% percent increase in
differences per basepair for every order of magnitude census size grows (95\%
confidence interval $[12\%, 14\%]$, adjusted $R^2 = 0.26$; see also the OLS fit
per-phyla, Supplementary Figure \ref{suppfig:figure-1-ave}).

Notably, this relationship has few outliers and is relatively homoscedastic.
This is in part because of the log-log scale, in contrast to previous work
\parencite{Nei1984-zi,Soule1976-he}; see Supplementary Figure
\ref{suppfig:figure-1-linear} for a version on a log-linear scale. However, it
is noteworthy that few taxa have diversity estimates below $10^{-3.5}$
differences per basepair. Those that do, lynx (\emph{Lynx Lynx}), wolverine
(\emph{Gulo gulo}), and Massasauga rattlesnake (\emph{Sistrurus catenatus})
face habitat fragmentation and declining population sizes. These three species
are all in the IUCN Red List, but are listed as least concern (though their
presence in the Red List indicates they are of conservation interest). In
Supplementary Material Section \Newnameref{supinfo:div-iucn}, I explore more
about the relationships between IUCN Red List status, diversity, and population
size.

@@

Finally, an important caveat is the increase in rate towards the tips could be
caused by measurement noise. Inspecting the lineage pairs that lead to this
increase in rate towards the tips indicates these represent plausible rate
shifts, e.g. between cosmopolitan and endemic sister species like
\emph{Drosophila simulans} and \emph{Drosophila sechellia}; however, ruling out
measurement noise entirely as an explanation would involve considering the
uncertainty of diversity and population size estimates.

@@

Previous work has found that the impact of linked selection increases with
$N_c$ (\cite{Corbett-Detig2015-gt}; see also Supplementary Figure
\ref{suppfig:corbett_detig}A), and it is often thought that this is driven by
higher rates of adaptive substitutions in larger populations, despite equivocal
evidence \parencite{Galtier2016-dq}. However, there is another mechanism by
which species with larger population sizes might experience a greater impact of
linked selection: recombinational map length, $L$, is known to correlate with
body mass \parencite{Burt1987-tq} and thus varies inversely with population
size. As this is a critical parameter that determines the genome-wide impact of
both hitchhiking and background selection, I examine the relationship between
recombination map length ($L$) and census population size ($N_c$) across taxa,
using available estimates of map lengths across species
\parencite{Stapley2017-fs,Corbett-Detig2015-gt}. I find a significant
non-linear relationship using phylogenetic mixed-effects models (Figure
\ref{fig:figure-3}A; see Methods: \Newnameref{sec:methods-pcm}). There is also
a correlation between map length and genome size (Supplementary Figure
\ref{supfig:genome_size_maplength}) and genome size and population size
(Supplementary Figure \ref{supfig:genome_size_popsize}). These findings are
consistent with both the hypothesis that non-adaptive processes increase genome
size in small-$N_e$ species \parencite{Lynch2003-bn} which in turn could
increase map lengths, as well as the hypothesis that map lengths are adaptively
longer to more efficiently select against deleterious alleles
\parencite{Roze2021-ya}. Overall, the negative relationship between map length
and census size indicates linked selection is expected to be stronger in short
map length, high-$N_c$ species.

@@

The relationship between recombination map length and the logarithm of
population size is non-linear and heteroscedastic, and was fit using a
lognormal phylogenetic mixed-effects model on the 130 species with complete
data. Since social insects have longer recombination map lengths
\parencite{Wilfert2007-dx}, social taxa were excluded when fitting this model.
All \texttt{Rhat} \parencite{Vehtari2019-po} values were below 1.01 and the
effective number of samples was over 1,000, consistent with good mixing;
details about the model are available in the GitHub repository
(\texttt{phylo\_mm\_lognormal.stan}). Continuous trait maps (Figure
\ref{fig:figure-2}A and Supplementary Material Figure \ref{suppfig:biconmap})
were created using \texttt{phytools} \parencite{Revell2012-zq}. Node-height
tests were implemented based on the methods in \texttt{Geiger}
\parencite{Pennell2014-nx,Harmon2008-tr}, and use robust regression to fit a
linear relationship between phylogenetic independent contrasts and branching
times.

@@

I would like to thank Andy Kern and Peter Ralph for helpful discussions and
supporting me during this work, and Graham Coop for inspiration and helpful
feedback during socially distanced nature walks at Yolo Basin. I thank Jessica
Stapley for kindly providing the recombination map length data, and Yaniv
Brandvain, Amy Collins, Doc Edge, Tyler Kent, Chuck Langley, Matt Osmond, Sally
Otto, Jeff Ross-Ibarra, Aaron Stern, Anastasia Teterina, Michael Turelli,
Margot Wood, and my Kern-Ralph labmates for helpful discussions. Sarah
Friedman, Katherine Corn, and Josef Uyeda provided very useful advice about
phylogenetic comparative methods; yet I take full responsibility for any
shortcomings of my analysis. Finally, I am indebted to helpful feedback from
Guy Sella, Matt Pennell, and two other anonymous reviewers. I would like to
also thank UO librarian Dean Walton for helping me track down some rather
difficult to find older papers. This work was supported by an NIH Grant
(1R01GM117241) awarded to Andrew Kern.

@@

\setcounter{table}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{equation}{0}
\renewcommand{\theequation}{S\arabic{equation}}
\setcounter{section}{0}
\renewcommand{\thesection}{S\arabic{section}}
\setcounter{figure}{0}
\renewcommand{\thefigure}{S\arabic{figure}}

@@

\begin{table}[ht]
  \resizebox{\textwidth}{!}{%
\centering
\begin{tabular}{lr|rr|rrrrr|r}
  &  & \multicolumn{2}{c}{\bf Bar-On et al.}  & \multicolumn{5}{c}{\bf Present study}  &  \\
  phylum & total species ($T$) & biomass ($B$) & prop. biomass & biomass ($b$) &  prop. biomass & num. species ($n$) & factor overrepresented &prop. total species ($f = \nicefrac{n}{T}$) & factor ($\nicefrac{b}{f B}$) \\
  \hline
  Arthropoda & $1.26 \times 10^{6}$ & 1.20 & 0.4635 & $2.78 \times 10^{-4}$ & 0.0100 &  68 & 0.02 & $5.41 \times 10^{-5}$ & 4.29 \\
  Chordata & $5.41 \times 10^{4}$ & 0.87 & 0.3357 & $2.70 \times 10^{-2}$ & 0.9719 &  68 & 2.90 & $1.26 \times 10^{-3}$ & 24.70 \\
  Annelida & $1.70 \times 10^{4}$ & 0.20 & 0.0772 & $1.23 \times 10^{-5}$ & 0.0004 &   3 & 0.01 & $1.76 \times 10^{-4}$ & 0.35 \\
  Mollusca & $9.54 \times 10^{4}$ & 0.20 & 0.0772 & $4.54 \times 10^{-4}$ & 0.0163 &  13 & 0.21 & $1.36 \times 10^{-4}$ & 16.60 \\
  Cnidaria & $1.60 \times 10^{4}$ & 0.10 & 0.0386 & $3.14 \times 10^{-5}$ & 0.0011 &   2 & 0.03 & $1.25 \times 10^{-4}$ & 2.51 \\
  Nematoda & $2.50 \times 10^{4}$ & 0.02 & 0.0077 & $3.89 \times 10^{-6}$ & 0.0001 &   1 & 0.02 & $4.00 \times 10^{-5}$ & 4.87 \\
\end{tabular}}
  \caption{How the total carbon biomass estimates by phylum from
    \textcite{Bar-On2018-kc} compare to the implied biomass estimates from this
    study. All biomass estimates are carbon biomass, and the proportions are of
    total biomass with respect to the study. The proportion of biomass in this
    study compared to the Bar-On et al. estimates \textcite{Bar-On2018-kc}
    indicates chordates are overrepresented and arthropods are underrepresented
    in the present study; the factor that each phylum is overrepresented is
    given in the eighth column. Total species by phylum estimates are from
    \textcite{Reaka-Kudla1996-fe,Nicol1969-qx,Zhang2013-zc,Chapman2009-fw}. The
    ratio column is the ratio of total biomass implied by the $N_c$ estimates
    of each species in a phylum to the actual biomass of that phylum.}
    \label{supp:biomass-tbl}
\end{table}

First, I compared the relative carbon biomass in this study to the relative
carbon biomass on earth per phylum. This shows that this study's sample is over
represents chordate biomass (by a factor of $\mathord{\sim}3$), and under represents
in arthropod biomass (by a factor of $0.02$) relative to the proportion of
carbon biomass of these phyla on earth (see column eight of Supplementary Table
\ref{supp:biomass-tbl}. Second, to check whether the carbon biomass per phylum
in the sample was broadly consistent with the total on earth by phylum ($B_S$
for phylum $S$), I calculated the expected sample biomass if species were
sampled randomly from the total species in a phylum, ($B_S \times
\nicefrac{n_S}{T_S}$, where $n_S$ is the total number of species in the sample
in phylum $S$, $T_S$ is the total number of species in phylum $S$ on earth).
The fraction of total species on earth included in the sample in this study is
depicted in Supplementary Figure \ref{suppfig:species_counts}.

Next, I look at the ratio of sample biomass per phylum, $b_S$ to this expected
biomass per phylum (Supplementary Table \ref{supp:biomass-tbl}). The
consistency is quite close for this rough approach and the non-random sample of
taxa included in this study. The carbon biomass estimates for chordates implied
by the census size estimates are $\mathord{\sim}25$-fold higher than expected, but is
well within reasonable expectations given that the chordate sample includes
many larger-bodied domesticated species (and is a biased sample in other ways).
Similarly, the implied arthropod carbon biomass is quite close to what one
would expect. Overall, these values indicate that the census size estimates
here do not lead to implied biomasses per phylum that are outside the range of
plausibility.

@@

As another consistency check, I looked at the rank order of mammals by biomass.
Whale species have the first and third highest biomass with 11.4 and 3.8
megatons of carbon biomass (for \emph{Balaenoptera bonaerensis} and
\emph{Eschrichtius robustus}, respectively). While this seems high, a recent
study shows that across whale species, pre-whaling carbon biomass was at the
tens of megatons level \parencite[Table 1 and Figure 1]{Pershing2010-fh}.
Given that my census size estimates represent populations at a macroecological
equilibrium, they would not reflect reduced density due to whaling or other
anthropogenic causes. Humans had the second largest biomass, followed by wolf
species (\emph{Canis lupus} and \emph{C. latrans}); as with whales, the
population sizes for wolf species represent pre-anthropogenic densities and are
overestimates compared to current population sizes as expected.

@@

\begin{table}[ht]
  \centering
  \begin{tabular}{rrrr}
    \hline
     & mean & 2.5 \% & 97.5 \% \\ 
 \hline
    $\beta_0$ & -2.80 & -3.20 & -2.50 \\ 
    $\beta_{LC}$ & -0.39 & -0.57 & -0.21 \\ 
    $\beta_{NT}$ & -0.22 & -0.83 & 0.39 \\ 
    $\beta_{VU}$ & -0.34 & -0.84 & 0.16 \\ 
    $\beta_{EN}$ & -0.40 & -0.73 & -0.07 \\ 
    $\beta_{CR}$ & -0.03 & -0.65 & 0.59 \\ 
    $\beta_{N_c}$ &  0.08 & 0.05 & 0.11 \\ 
    \hline
  \end{tabular}
  \caption{TODO The regression estimates of full IUCN Red List population size model
    for diversity, $\log_{10}(\pi) = \beta_0 + \beta_{LC} LC + \beta_{NT} NT +
    \beta_{VU} VU + \beta_{EN} EN + \beta_{CR} CR + \beta_{N_c} \log_{10}(N_c)$;
    $df = 165$.  Using AIC to compare this full model to a reduced model of
    $\log_{10}(\pi) = \beta_0 + \beta_{N_c} \log_{10}(N_c)$, $AIC_\text{full} =
  203.99$, $AIC_\text{reduced} =212.46$. }
\end{table}
\label{supptable:tbl-1}

@@

  \caption{The appendix table of \textcite{Damuth1987-sg}; the color indicates
    Damuth's original group labels. The dashed line was estimated using a
    lognormal regression model in Stan. References to each measurement are
  available in \textcite{Damuth1987-sg}.}

  \label{suppfig:damuth}
\end{figure}

>>>>

\title{Natural Selection is Unlikely to Explain Why Species Get a Thin Slice of $\pi$}
%\title{Why do species get a thin slice of $\pi$? Revisiting Lewontin's Paradox of Variation}

@@

A major impediment in quantifying Lewontin's Paradox has been estimating census
population sizes across many taxa, especially for extremely abundant,
cosmopolitan species that define the upper limit of ranges. Previous work has
surveyed the literature for census size estimates
\parencite{Nei1984-zi,Soule1976-he,Frankham1996-yb}, or used range, body size,
or qualitative categories as proxies for census size
\parencite{Corbett-Detig2015-gt,Leffler2012-zj}. To quantify the relationship
between genomic estimates of diversity and census population sizes, I first
approximate census population sizes for 172 metazoan taxa (Figure
\ref{fig:figure-nc}). My approach predicts population densities from body sizes
using a previously-observed linear relationship that holds across metazoans
(Figure 1-figure supplement 1 \ref{suppfig:damuth};
\cite{Damuth1981-st,Damuth1987-sg}). Then, from geographic occurrence data, I
estimate range sizes.  Finally, I estimate population size as the product of
these predicted densities and range estimates (see Methods:
\Newnameref{sec:methods-popsize}). Note that the relationship between
population density and body size is driven by energy budgets, and thus reflects
macroecological equilibria \parencite{Damuth1987-sg}. Consequently, population
sizes are underestimated for taxa like humans and their domesticated species,
and overestimated for species with anthropogenically reduced densities or
fragmented ranges. For example, the population size of \emph{Lynx lynx} is
likely around 50,000 \parencite{Iucn2020-ap} which is around two orders of
magnitude smaller than my estimate. Additionally, the range size estimates do
not consider whether an area has unsuitable habitat, and thus may be
overestimated for species with particular niches or patchy habitats. While
these methods to estimate census size are crude and approximate, they can be
efficiently calculated for numerous taxa and are sufficient to estimate the
scale of Lewontin's Paradox (see Supplementary Information
\Newnameref{supp:popsize-valid} for more on validation based on biomass and
other approaches).

@@

To determine which ecological or evolutionary processes could decouple
diversity from census population size, we first need to quantify this
relationship across a wide variety of taxa. Previous work has found there is a
significant relationship between heterozygosity and the logarithm of population
size or range size, but these studies relied on heterozygosity measured from
allozyme data
\parencite{Soule1976-he,Frankham1996-yb,Nei1984-zi,Leffler2012-zj}. Here, I
confirm these findings using pairwise diversity estimates from genomic sequence
data and the estimated census sizes (Figure \ref{fig:figure-1}). The pairwise
diversity estimates are from three sources: \textcite{Leffler2012-zj},
\textcite{Corbett-Detig2015-gt}, and \textcite{Romiguier2014-bp}, and are
predominantly from either synonymous or non-coding DNA (see Methods:
\Newnameref{sec:methods-diversity}). Overall, an ordinary least squares (OLS)
relationship on a log-log scale fits the data well (Figure \ref{fig:figure-1}).
The OLS slope estimate is significant and implies a 13\% percent increase in
differences per basepair for every order of magnitude census size grows (95\%
confidence interval $[12\%, 14\%]$, adjusted $R^2 = 0.26$; see also the OLS fit
per-phyla, Figure 1-figure supplement \ref{suppfig:figure-1-ave}).

Notably, this relationship has few outliers and is relatively homoscedastic.
This is in part because of the log-log scale, in contrast to previous work
\parencite{Nei1984-zi,Soule1976-he}; see Figure 1-supplement figure 2 for a
version on a log-linear scale. However, it is noteworthy that few taxa have
diversity estimates below $10^{-3.5}$ differences per basepair. Those that do,
lynx (\emph{Lynx Lynx}), wolverine (\emph{Gulo gulo}), and Massasauga
rattlesnake (\emph{Sistrurus catenatus}) face habitat fragmentation and
declining population sizes. These three species are all in the IUCN Red List,
but are listed as least concern (though their presence in the Red List
indicates they are of conservation interest). In Supplementary Information
Section \Newnameref{supinfo:div-iucn}, I explore more about the relationships
between IUCN Red List status, diversity, and population size.

@@

Finally, an important caveat is the increase in rate towards the tips could be
caused by measurement noise, or possibly uncertainty or bias in the divergence
time estimates deep in the tree. Inspecting the lineage pairs that lead to this
increase in rate towards the tips indicates these represent plausible rate
shifts, e.g.  between cosmopolitan and endemic sister species like
\emph{Drosophila simulans} and \emph{Drosophila sechellia}; however, ruling out
measurement noise entirely as an explanation would involve considering the
uncertainty of diversity and population size estimates.

@@

Previous work has found that the impact of linked selection increases with
$N_c$ (\cite{Corbett-Detig2015-gt}; see also Supplementary Figure
\ref{suppfig:corbett_detig}A), and it is often thought that this is driven by
higher rates of adaptive substitutions in larger populations
\parencite{Ohta1992-yi}, despite equivocal evidence \parencite{Galtier2016-dq}.
However, there is another mechanism by which species with larger population
sizes might experience a greater impact of linked selection: recombinational
map length, $L$, is known to correlate with body mass \parencite{Burt1987-tq}
and thus varies inversely with population size. As this is a critical parameter
that determines the genome-wide impact of both hitchhiking and background
selection, I examine the relationship between recombination map length ($L$)
and census population size ($N_c$) across taxa, using available estimates of
map lengths across species \parencite{Stapley2017-fs,Corbett-Detig2015-gt}. I
find a significant non-linear relationship using phylogenetic mixed-effects
models (Figure \ref{fig:figure-3}A; see Methods: \Newnameref{sec:methods-pcm}).
There is also a correlation between map length and genome size (Supplementary
Figure \ref{supfig:genome_size_maplength}) and genome size and population size
(Supplementary Figure \ref{supfig:genome_size_popsize}). These findings are
consistent with both the hypothesis that non-adaptive processes increase genome
size in small-$N_e$ species \parencite{Lynch2003-bn} which in turn could
increase map lengths, as well as the hypothesis that map lengths are adaptively
longer to more efficiently select against deleterious alleles
\parencite{Roze2021-ya}. Overall, the negative relationship between map length
and census size indicates linked selection is expected to be stronger in short
map length, high-$N_c$ species.

@@

The relationship between recombination map length and the logarithm of
population size is non-linear and heteroscedastic, and was fit using a
lognormal phylogenetic mixed-effects model on the 130 species with complete
data. Since social insects have longer recombination map lengths
\parencite{Wilfert2007-dx}, social taxa were excluded when fitting this model.
All \texttt{Rhat} \parencite{Vehtari2019-po} values were below 1.01 and the
effective number of samples was over 1,000, consistent with good mixing;
details about the model are available in the GitHub repository
(\texttt{phylo\_mm\_lognormal.stan}). Continuous trait maps (Figure
\ref{fig:figure-2}A and Supplementary Information Figure \ref{suppfig:biconmap})
were created using \texttt{phytools} \parencite{Revell2012-zq}. Node-height
tests were implemented based on the methods in \texttt{Geiger}
\parencite{Pennell2014-nx,Harmon2008-tr}, and use robust regression to fit a
linear relationship between phylogenetic independent contrasts and branching
times.

@@

I would like to thank Andy Kern and Peter Ralph for helpful discussions and
supporting me during this work, and Graham Coop for inspiration and helpful
feedback during socially distanced nature walks at Yolo Basin. I thank Jessica
Stapley for kindly providing the recombination map length data, and Yaniv
Brandvain, Amy Collins, Doc Edge, Tyler Kent, Chuck Langley, Matt Osmond, Sally
Otto, Jeff Ross-Ibarra, Aaron Stern, Anastasia Teterina, Michael Turelli,
Margot Wood, and my Kern-Ralph labmates for helpful discussions. Sarah
Friedman, Katherine Corn, and Josef Uyeda provided very useful advice about
phylogenetic comparative methods; yet I take full responsibility for any
shortcomings of my analysis. Finally, I am indebted to Guy Sella, Matt Pennell,
and two other anonymous reviewers for helpful feedback. I would like to also
thank UO librarian Dean Walton for helping me track down some rather difficult
to find older papers.  This work was supported by an NIH Grant (1R01GM117241)
awarded to Andrew Kern.

@@

\setcounter{table}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{equation}{0}
\renewcommand{\theequation}{S\arabic{equation}}
\setcounter{section}{0}
\renewcommand{\thesection}{S\arabic{section}}

@@

\begin{table}[ht]
  \resizebox{\textwidth}{!}{%
\centering
\begin{tabular}{lr|rr|rrrrr|r}
  &  & \multicolumn{2}{c}{\bf Bar-On et al.}  & \multicolumn{5}{c}{\bf Present study}  &  \\
  phylum & total species ($T$) & biomass ($B$) & prop. biomass & biomass ($b$) &  prop. biomass & num. species ($n$) & factor overrepresented &prop. total species ($f = \nicefrac{n}{T}$) & factor ($\nicefrac{b}{f B}$) \\
  \hline
  Arthropoda & $1.26 \times 10^{6}$ & 1.20 & 0.4635 & $2.80 \times 10^{-4}$ & 0.0102 &  68 & 0.02 & $5.41 \times 10^{-5}$ & 4.31 \\ 
  Chordata & $5.41 \times 10^{4}$ & 0.87 & 0.3357 & $2.67 \times 10^{-2}$ & 0.9715 &  68 & 2.89 & $1.26 \times 10^{-3}$ & 24.40 \\ 
  Annelida & $1.70 \times 10^{4}$ & 0.20 & 0.0772 & $1.23 \times 10^{-5}$ & 0.0004 &   3 & 0.01 & $1.76 \times 10^{-4}$ & 0.35 \\ 
  Mollusca & $9.54 \times 10^{4}$ & 0.20 & 0.0772 & $4.56 \times 10^{-4}$ & 0.0166 &  13 & 0.21 & $1.36 \times 10^{-4}$ & 16.70 \\ 
  Cnidaria & $1.60 \times 10^{4}$ & 0.10 & 0.0386 & $3.07 \times 10^{-5}$ & 0.0011 &   2 & 0.03 & $1.25 \times 10^{-4}$ & 2.45 \\ 
  Nematoda & $2.50 \times 10^{4}$ & 0.02 & 0.0077 & $4.03 \times 10^{-6}$ & 0.0001 &   1 & 0.02 & $4.00 \times 10^{-5}$ & 5.03 \\
\end{tabular}}
\caption{How the total carbon biomass estimates by phylum from
  \textcite{Bar-On2018-kc} compare to the implied biomass estimates from this
  study. All biomass estimates are carbon biomass, and the proportions are of
  total biomass with respect to the study. The proportion of biomass in this
  study compared to the Bar-On et al. estimates \textcite{Bar-On2018-kc}
  indicates chordates are overrepresented and arthropods are underrepresented
  in the present study; the factor that each phylum is overrepresented is
  given in the eighth column. Total species by phylum estimates are from
  \textcite{Reaka-Kudla1996-fe,Nicol1969-qx,Zhang2013-zc,Chapman2009-fw}. The
  ratio column is the ratio of total biomass implied by the $N_c$ estimates
of each species in a phylum to the actual biomass of that phylum.}
\label{supp:biomass-tbl}
\end{table}

First, I compared the relative carbon biomass in this study to the relative
carbon biomass on earth per phylum. This shows that this study's sample over
represents chordate biomass (by a factor of $\mathord{\sim}3$), and under
represents in arthropod biomass (by a factor of $0.02$) relative to the
proportion of carbon biomass of these phyla on earth (see column eight of
Supplementary Table \ref{supp:biomass-tbl}. Second, to check whether the carbon
biomass per phylum in the sample was broadly consistent with the total on earth
by phylum ($B_S$ for phylum $S$), I calculated the expected sample biomass if
species were sampled randomly from the total species in a phylum, ($B_S \times
\nicefrac{n_S}{T_S}$, where $n_S$ is the total number of species in the sample
in phylum $S$, $T_S$ is the total number of species in phylum $S$ on earth).
The fraction of total species on earth included in the sample in this study is
depicted in Supplementary Figure \ref{suppfig:species_counts}.

Next, I look at the ratio of sample biomass per phylum, $b_S$ to this expected
biomass per phylum (Supplementary Table \ref{supp:biomass-tbl}). The
consistency is quite close for this rough approach and the non-random sample of
taxa included in this study. The carbon biomass estimates for chordates implied
by the census size estimates are $\mathord{\sim}24$-fold higher than expected, but is
well within reasonable expectations given that the chordate sample includes
many larger-bodied domesticated species (and is a biased sample in other ways).
Similarly, the implied arthropod carbon biomass is quite close to what one
would expect. Overall, these values indicate that the census size estimates
here do not lead to implied biomasses per phylum that are outside the range of
plausibility.

@@

As another consistency check, I looked at the rank order of mammals by biomass.
Whale species have the first and third highest biomass with 11.4 and 3.9
megatons of carbon biomass (for \emph{Balaenoptera bonaerensis} and
\emph{Eschrichtius robustus}, respectively). While this seems high, a recent
study shows that across whale species, pre-whaling carbon biomass was at the
tens of megatons level \parencite[Table 1 and Figure 1]{Pershing2010-fh}.
Given that my census size estimates represent populations at a macroecological
equilibrium, they would not reflect reduced density due to whaling or other
anthropogenic causes. Humans had the second largest biomass, followed by wolf
species (\emph{Canis lupus} and \emph{C. latrans}); as with whales, the
population sizes for wolf species represent pre-anthropogenic densities and are
overestimates compared to current population sizes, as expected.

@@

\begin{table}[ht]
  \centering
  \begin{tabular}{rrrr}
    \hline
     & mean & 2.5 \% & 97.5 \% \\ 
 \hline
    $\beta_0$ & -2.80 & -3.20 & -2.50 \\ 
    $\beta_{LC}$ & -0.39 & -0.57 & -0.21 \\ 
    $\beta_{NT}$ & -0.22 & -0.83 & 0.39 \\ 
    $\beta_{VU}$ & -0.34 & -0.84 & 0.16 \\ 
    $\beta_{EN}$ & -0.40 & -0.73 & -0.07 \\ 
    $\beta_{CR}$ & -0.03 & -0.65 & 0.59 \\ 
    $\beta_{N_c}$ &  0.08 & 0.05 & 0.11 \\ 
    \hline
  \end{tabular}
  \caption{The regression estimates of full IUCN Red List population size model
    for diversity, $\log_{10}(\pi) = \beta_0 + \beta_{LC} LC + \beta_{NT} NT +
    \beta_{VU} VU + \beta_{EN} EN + \beta_{CR} CR + \beta_{N_c} \log_{10}(N_c)$;
    $df = 165$.  Using AIC to compare this full model to a reduced model of
    $\log_{10}(\pi) = \beta_0 + \beta_{N_c} \log_{10}(N_c)$, $AIC_\text{full} =
 204.9$, $AIC_\text{reduced} = 216.4$. }
\end{table}
\label{supptable:tbl-1}

\section{Figure Supplement}

\setcounter{figure}{0}
\renewcommand{\thefigure}{}
\renewcommand{\figurename}{Figure 1-figure supplement 1}

  \caption{The appendix table of \textcite{Damuth1987-sg}; the color indicates
    Damuth's original group labels. The dashed line was estimated using a
    lognormal regression model in Stan. References to each measurement are
  available in \textcite{Damuth1987-sg}.}
  \label{suppfig:damuth}
\end{figure}

@@
